Clinical Trial Design Using A Stopped Negative Binomial Distribution

### 2017-02-27 stat-and-prob-letters/snb-paper.tex
yeah.

\documentclass[review]{elsarticle}

@@

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amsthm}
%\usepackage{url}
\usepackage{subfig}
\DeclareMathOperator*{\argmin}{argmin}
\newcommand*{\argminl}{\argmin\limits}
\newtheorem{prop}{Proposition}
\newtheorem{corollary}{Corollary}

@@

\journal{Statistics and Probability Letters}

@@

%% `Elsevier LaTeX' style
\bibliographystyle{elsarticle-num}
%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{frontmatter}

\title{A Stopped Negative Binomial Distribution}
\tnotetext[mytitlenote]{This research was supported by grants R01CA131301, 
R01CA157749, R01CA148996, R01CA168733, and PC50CA196530 awarded by the 
National Cancer Institute along with support from the Yale Comprehensive Cancer 
Center and the Yale Center for Outcomes Research. We would also like 
to thank Rick Landin at Ignyta Inc. for his suggestions.}

@@

%% or include affiliations in footnotes:
\author{Michelle DeVeaux}
\ead{michelle.deveaux@yale.edu}

\author{Michael J. Kane$^*$\footnote{* Corresponding author}}
\ead{michael.kane@yale.edu}

\author{Daniel Zelterman}
\ead{daniel.zelterman@yale.edu}

\address{Department of Biostatistics\\ School of Epidemiology and Public Health\\ Yale University, New Haven, CT}

@@

\begin{keyword}
discrete distribution\sep curtailed sampling
\end{keyword}

\end{frontmatter}

@@

Consider a prototypical early phase, single-arm clinical trial in which 
17 patients
are enrolled and treated. The binomial probability of a patient 
responding to treatment is $p=0.2$ 
under the null hypothesis that the treatment is not effective.
If seven or more patients out of these 17 respond to the treatment then we 
reject this hypothesis and the treatment is deemed successful at 
a significance level of $0.1$.  If fewer than seven respond then the null 
hypothesis is not 
rejected and the treatment is deemed ineffective.

@@

Let us assume each patient outcome can be modeled as an independent,
identically distributed Bernoulli($p$) random variable. The trial is realized
as a sequence of Bernoulli samples stopping when either a
specified number of responders or non-responders has been reached.

\begin{figure}[bp!]
\centering
\includegraphics[width=0.9\textwidth]{KanePlot.pdf}
\caption{
A hypothetical realization of the trial.
}
\label{fig:kane_viz}
\end{figure}

A hypothetical sample path is illustrated in Fig.~\ref{fig:kane_viz}.
The vertical axis denotes the number of
successful outcomes. The horizontal axis counts the number of patients 
enrolled. The horizontal and vertical boundaries represent
endpoints for the trial. In this case, a seventh response was reached on
the $15^{\text{th}}$ enrollment.
Since the success boundary is reached, we say the treatment succeeds.

@@

\begin{figure}[bp!]
\centering
\subfloat[enrollment distribution]{\includegraphics[width=0.45\textwidth]{snb-first-plot.pdf}}
\hfill
\subfloat[mean and variance varying $p$]{\includegraphics[width=0.45\textwidth]{mean-and-variance.pdf}}
\caption{
The distribution of the sample size in a trial that stops after seven 
patients respond to treatment
or 11 patients do not. Panel (a) shows the distribution when
$p = 0.2$. The mean and variance of the distribution varying $p$ 
between zero and one are shown in panel (b).
}
\label{fig:exp-and-var}
\end{figure}

More generally the distribution of the number of trial enrollments is shown in 
Fig.~\ref{fig:exp-and-var} (a). There is relatively little probability mass
for values of sample size equal to seven through 10 since $p$ is small and it 
is unlikely the treatment will succeed quickly.
%The probability mass is concentrated at the 
%$11^{\text{th}}$ and $12^{\text{th}}$ step since $p$ 
%is small and for most realizations reach 11 
%non-responders before reaching 2 responders.
Fig.~\ref{fig:exp-and-var} (b) shows the expected value and variance for the
number of trial enrollments varying $p$ between zero and one. When $p$ is
small then the treatment
is more likely to fail shortly after the $11^{\text{th}}$ enrollment.
When $p$ is large then the treatment is more likely to succeed and the 
number of enrollees approaches seven from above.

@@

Let $\,b_1, b_2, \ldots \,$ denote a sequence of independent, identically
distributed, Bernoulli random variables with $\mathbb{P}[b_i=1]=p$ and
$\mathbb{P}[b_i = 0] = 1-p$, for
probability parameter $0\leq p \leq 1$. In the clinical trial setting
$\,b_i = 1$ corresponds to a patient responding to treatment.  
Let $s$ and $t$ be positive integers.  Define the SNB random
variable $Y$ as the smallest
integer value such that $\,\{b_1, \ldots , b_Y\}\,$ contains {\em either}
$\,s\,$ responders {\em or} $\,t\,$ non-responders. That is, the SNB 
distribution of $Y$ is the smallest integer such that either
$\sum_i^Y b_i = s$ or $\sum_i^Y 1-b_i = t$.

The distribution of $\,Y\,$ has support on integer values in the range
\begin{equation*}               
     \min(s,t) \leq \; Y \;\leq s+t-1  \label{range.y.eq}.
\end{equation*}
The probability mass function is
\begin{equation} \label{eqn:pmf}
\mathbb{P} [Y=k] = S(k, p, s) \ I_{\{s \leq k \leq s+t-1\}} + 
  S(k, 1-p, t) \ I_{\{ t \leq k \leq s+t-1 \}}
\end{equation}
where $I_{\{f\}}$ is the {\em indicator function}, taking the value 
of one if $f$ is true and zero otherwise, and
\begin{equation} \label{eqn:N}
S(k, p, s) = {k-1 \choose s-1} p^s (1-p)^{k-s} 
\end{equation}
is the negative binomial probability mass.

@@

The probability a given realization of $\mathbf{X}$ reaches $s$ at
the $k$th outcome is the probability that, at time $k-1$, there are $s-1$
successful outcomes and $k-s$ unsuccessful outcomes multiplied by
the probability of a final success at time $k$. This expression is given
in (\ref{eqn:N}). 
Similarly, the probability a given realization reaches $k-t$
is the probability that, at outcome $k-1$, there are $k-t$ successful outcomes
and $t-1$ unsuccessful outcomes multiplied by the probability of a final
unsuccessful outcome at time $k$.

@@

To show that (\ref{eqn:pmf}) sums to one, define
%\begin{align} \label{eqn:sum_proof}
%R &= \sum_{k=s}^{s+t-1} S(k, p, s) + \sum_{k=t}^{s+t-1} S(k, 1-p, t) \\
%  &= \sum_{k=s}^{s+t-1} {k-1 \choose s-1} p^s (1-p)^{k-s} + \sum_{k=t}^{s+t-1} {k-1 \choose k-t} p^{k-t} (1-p)^t
%\end{align}
\begin{equation*} 
R = \sum_{k=s}^{s+t-1} S(k, p, s) + \sum_{k=t}^{s+t-1} S(k, 1-p, t).
\end{equation*}
If we substitute $i=k-s$ in the first summation and $j=k-t$ in the second then
$R$ can be written as the cumulative distribution function (CDF) of two
negative binomial distributions:
\begin{equation} \label{eqn:transformed_sum}
R = \sum_{i=0}^{t-1} {i+s-1 \choose i} p^s (1-p)^i \; + \;
\sum_{j=0}^{s-1} {j+t-1 \choose j} p^j (1-p)^t.
\end{equation}

The CDF of the negative binomial is commonly expressed in terms of the
{\em regularized incomplete beta function} \citep{Olver2010}. Using standard
notation, the first summation in (\ref{eqn:transformed_sum})
is written as $\mathcal{I}_{1-p}(t, s)$.
This function satisfies 
$\mathcal{I}_p(s, t) = 1-\mathcal{I}_{1-p}(t, s)$ \citep{Uppuluri1967}. Then
\begin{align*}
R = \sum_{i=0}^{t-1} &{i+s-1 \choose i} p^s (1-p)^i +
\sum_{j=0}^{s-1}  {j+t-1 \choose j} p^j  (1-p)^t \\
   &= 1-\mathcal{I}_p(s, t) + 1 - \mathcal{I}_{1-p}(t, s) \\
   &= 1. 
\end{align*}
This completes the proof that (\ref{eqn:pmf}) is the distribution of the
stopping time and is a valid probability mass function.
\end{proof}

@@

\begin{figure}[p!]
\begin{center}
\includegraphics[width=0.8\textwidth]{shapes.pdf}
\end{center}
\caption{Different shapes of the SNB distribution with parameters 
($p$, $s$, $t$), as given. Black indicates mass contributed by reaching
$s$ responders before $t$ non-responders. Grey indicates
mass contributed by reaching $t$ non-responders first. \label{shapes.fig}}
\end{figure}

@@

The SNB generalizes both the minimum (riff-shuffle) and maximum negative
binomial distributions up to a translation of the support.
For the special case of $\,s=t,$ the distribution of $\,Y\,$ is the
riff-shuffle, or minimum negative binomial 
distribution~\citep{Uppuluri1967,Johnson2005}.
The maximum negative binomial \cite{Johnson2005,Zhang2000,Zelterman2005} is 
the smallest number of outcomes necessary to 
observe at least $s$ responders {\em and} $s$ non-responders and is equivalent
to a translated version of the riff-shuffle.

@@

%There is a close connection between the tail probabilities of the SNB and the 
%binomial distributions.
%The probability of reaching the success endpoint in an 
%SNB($p$, $s$, $t$) random variable is 
%equal to the probability of at least $s$ successes in a binomial distributed 
%random variable with size $s+t-1$ and success probability $p$. 
%Likewise, the probability of reaching the failure endpoint is equal
%to the probability of at most $s-1$ successes in a binomial distribution. 
There is an equivalence between the probability of reaching an endpoint 
in the SNB model and the tail probability of the binomial distribution.
That is, the probability that the number of responders is at least $s$ in the 
binomial model is the same as the probability the treatment succeeds 
(reaches $s$) in the SNB 
model.
\begin{prop} \label{binomial_tail}
Let $Y$ be distributed as SNB($p$, $s$, $t$) and let $X_Y$ correspond
to the number of responders at the end of the trial. Let 
$B$ be distributed binomial with size $n=s+t-1$ and response probability
$p$. Then
\begin{equation}
\mathbb{P}[B \geq s] = \mathbb{P} [X_Y = s].
\end{equation}
\end{prop}
\begin{proof}
The binomial tail probability is
\begin{equation*}
\mathbb{P}[B \geq s] = 1 - \mathcal{I}_{1-p}(s, t)
\end{equation*}
%\begin{align*}
%\mathbb{P}[B \geq s] &= \sum_{k=s}^{s+t-1} {n \choose k} p^k (1-p)^{n-k} \\
%  &= 1 - \sum_{k=0}^{s-1} {n \choose k} p^k (1-p)^{n-k} \\
%  &= 1 - \mathcal{I}_{1-p}.
%\end{align*}
where $\mathcal{I}_{1-p}(s, t)$ is the {\em regularized incomplete beta} 
function. The corresponding success probability is
\begin{equation} \label{eqn:success}
\mathbb{P} [X_Y = s] 
  = \sum_{k=s}^{s+t-1} {k-1 \choose s-1} p^s (1-p)^{k-s}.
\end{equation}
Let $i=k-s$. Since
\begin{equation*}
{i+s-1 \choose s-1} = {i+s-1 \choose i},
\end{equation*}
the summation in (\ref{eqn:success}) can be rewritten as
\begin{align*}
\mathbb{P} [X_Y = s] &= \sum_{i=0}^{t-1} {i+s-1 \choose i} p^s (1-p)^i\\
  &= 1 - \mathcal{I}_{1-p}(t, s)
\end{align*}
completing the proof.
\end{proof}

@@

The moment generating function for the SNB is calculated in a manner similar to 
that of two negative binomial distributions. 
\begin{prop} Let $Y$ be distributed SNB with parameters $p$, $s$, and $t$.
Then the moment generating function (MGF) of $Y$ is
\begin{equation} \label{eqn:mgf}
\mathbb{E}~e^{xY} = \left(\frac{p e^x}{1 - qe^x}\right)^s 
  \mathcal{I}_{1-qe^x} (s, t) + \left(\frac{qe^x}{1-pe^x}\right)^t 
  \mathcal{I}_{1-pe^x}(t, s)
\end{equation}
for $q = 1-p$ and is defined for 
$x < \min \left\{\log(1/p), \log(1/q) \right\}$.
\end{prop}
\begin{proof}
The MGF of the SNB is:
\begin{equation*}
\mathbb{E}~e^{xY} = \sum_{k=s}^{s+t-1} {k-1 \choose s-1} p^s q^{k-s} e^{kx} 
  + \sum_{k=t}^{s+t-1} {k-1 \choose t-1} p^{k-t} q^t e^{kx}
\end{equation*}
and can be rewritten as:
\begin{equation} \label{eqn:first_sum}
\mathbb{E}~e^{xY} = \sum_{k=s}^{s+t-1}{k-1 \choose s-1} (pe^x)^{s} (qe^x)^{k-s} 
  + \sum_{k=t}^{s+t-1}{k-1 \choose t-1} (qe^x)^t (pe^x)^{k-t}.
\end{equation}
The first summation in (\ref{eqn:first_sum}) satisfies
\begin{align*}
\sum_{k=s}^{s+t-1}{k-1 \choose s-1} (pe)^{sx} (qe^x)^{k-s} &= 
  \left(\frac{pe^x}{1 - qe^x}\right)^s \ \ \sum_{k=s}^{s+t-1} {k-1 \choose s-1} 
    (qe^x)^{k-s} (1-qe^x)^s \\
  &= \left(\frac{pe^x}{1 - qe^x}\right)^s \mathcal{I}_{1-qe^x}(s, t).
\end{align*}
Since the regularized incomplete beta function's subscript parameter 
has support on zero 
to one, we have $0 \leq qe^x < 1$. This implies
$x < -\log(q)$.
A similar expression can be derived for the second summation in 
(\ref{eqn:first_sum}) and results in
the constraint $x < -\log(p)$.
\end{proof}

The SNB's ability to approximate the geometric, normal, gamma, and Poisson
distributions follow from it generalizing the negative binomial distribution. 
\begin{prop}
The MGF of the SNB converges to that of the negative binomial when either
$s$ or $t$ gets large. That is
\begin{equation*}
\mathbb{E} e^{xY} \rightarrow \left( \frac{pe^x}{1-qe^x} \right)^s
\end{equation*}
as $t \rightarrow \infty$. The analogous result holds when 
$s \rightarrow \infty$.
\end{prop}
\begin{proof}
The second regularized incomplete beta term in (\ref{eqn:mgf}) can be written
in terms of a cumulative binomial distribution
\begin{equation*}
\mathcal{I}_{1-pe^x}(t, s) = \mathbb{P}\left[ B \leq s-1 \right]
\end{equation*}
where $B$ is distributed as
Binomial($t-k$, $pe^x$). By Chebychev's inequality %\cite{Hoeffding1963}
it follows that
\begin{equation} \label{eqn:hoeffding}
\mathbb{P}\left[ B \leq s-1 \right] \leq 
  \frac{ (t-k) pe^x (1-pe^x) }{ \left(s - (t-k)pe^x\right)^2 }
\end{equation}
As $t$ gets large $\mathcal{I}_{1-pe^x}(t, s)$ goes to zero
Likewise, $\mathcal{I}_{1-qe^x}(s, t)$ goes
to one. The proof follows by realizing 
\begin{equation*}
0 < \frac{qe^x}{1-pe^x} < 1
\end{equation*}
over the support of $x$.
\end{proof}

When $s=1$ (and $t$ is still large) the SNB's MGF is approximately 
the same as that of the
geometric distribution. The negative
binomial can therefore be seen as a sum of i.i.d. geometric distributions.
For an appropriately large number of samples, the central limit theorem
yields a normal approximation.

Drawing connections to the gamma and Poisson distributions are more complicated.
However a connection to the gamma distribution well-studied problem in the 
literature (see \cite{Best1974,Ord1968,Guenther1972} for examples). 
A connection to the Poisson appears in \cite{Anscombe1950} where
it is shown that if the mean of a Poisson is proportional
to a chi-square distribution with $2k$ degrees of freedom
then the negative binomial is obtained. 
Both of these approximations work by equating cumulants and then showing 
that differences between the cumulant generating
functions converge to zero.

The lower-half normal distribution can be approximated by setting $s=t$
for appropriately large $s$ and $t$ and $p = 0.5$. In this
case, the SNB can be viewed as identical, negative binomials
approximating a normal and truncated at the median.

@@

Let us consider the case where the rate parameter is distributed 
as Beta($\alpha$, $\beta$) and denoted $P$.
The prior times the likelihood is given by the function
\begin{align} \label{eqn:ptl}
f_P(p, \alpha, \beta ) f_{Y|P}(p, k, s, t) &= 
  \frac{ {k-1 \choose s-1} }{B(\alpha, \beta)} p^{\alpha +s -1} 
    (1-p)^{k+\beta-s-1} + \\
  & \frac{ {k-1 \choose t-1} }{B(\alpha, \beta)} p^{k+\alpha -t -1} 
    (1-p)^{\beta+t-1} \nonumber
\end{align}
where $0 \leq p \leq 1$, $s \leq k \leq s+k-1$, and $t \leq k \leq s+k-1$.
This result can be found directly by multiplying the probability mass function
in (\ref{eqn:pmf}) by the density of the Beta distribution.

>>>>

%\documentclass[review]{elsarticle}
\documentclass[12pt]{article}

@@

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{setspace}
\doublespacing
%\usepackage{url}
\usepackage{subfig}
\DeclareMathOperator*{\argmin}{argmin}
\newcommand*{\argminl}{\argmin\limits}
\newtheorem{prop}{Proposition}
\newtheorem{corollary}{Corollary}

@@

%\journal{Statistics and Probability Letters}

@@

%% `Elsevier LaTeX' style
%\bibliographystyle{elsarticle-num}
\bibliographystyle{plain}
%%%%%%%%%%%%%%%%%%%%%%%

@@

%\begin{frontmatter}

\title{Clinical Trial Design Using A Stopped Negative Binomial Distribution}
\maketitle
%\tnotetext[mytitlenote]{This research was supported by grants R01CA131301, 
%R01CA157749, R01CA148996, R01CA168733, and PC50CA196530 awarded by the 
%National Cancer Institute along with support from the Yale Comprehensive Cancer 
%Center and the Yale Center for Outcomes Research. We would also like 
%to thank Rick Landin at Ignyta Inc. for his suggestions.}

@@

%% or include affiliations in footnotes:
%\author{Michelle DeVeaux}
%\ead{michelle.deveaux@yale.edu}

%\author{Michael J. Kane$^*$\footnote{* Corresponding author}}
%\ead{michael.kane@yale.edu}

%\author{Daniel Zelterman}
%\ead{daniel.zelterman@yale.edu}

%\address{Department of Biostatistics\\ School of Epidemiology and Public Health\\ Yale University, New Haven, CT}

@@

%\begin{keyword}
%discrete distribution\sep curtailed sampling
%\end{keyword}

%\end{frontmatter}

@@

Consider a prototypical early phase, single-arm clinical trial in which 
17 patients
are enrolled and treated. Suppose the Bernoulli probability of a patient 
responding to treatment is $p=0.2$ 
under the null hypothesis that the treatment is not effective.
If seven or more patients out of these 17 respond to the treatment then we 
reject this hypothesis and the treatment is deemed successful at 
a significance level of $0.1$.  If fewer than seven respond then the null 
hypothesis is not 
rejected and the treatment is deemed ineffective. The trial is 
realized as a sequence of Bernoulli($p$) samples stopping when either
the specified number of responders or non-responders is reached.

@@

\begin{figure}[bp!]
\centering
\includegraphics[width=\textwidth]{KanePlot.pdf}
\caption{
A hypothetical realization of the trial.
}
\label{fig:kane_viz}
\end{figure}

A hypothetical sample path is illustrated in Fig.~\ref{fig:kane_viz}.
The vertical axis denotes the number of
successful outcomes. The horizontal axis counts the number of patients 
enrolled. The horizontal and tilted vertical boundaries represent
endpoints for the trial. In this case, a seventh response was reached on
the $15^{\text{th}}$ enrollment.
Since the success boundary is reached, we say the treatment succeeds.

@@

\begin{figure}[bp!]
\centering
\includegraphics[width=\textwidth]{snb-first-plot.pdf}
\caption{
The distribution of the sample size in a trial that stops after seven patients
respond to treatment or 11 do not when $p=0.2$.
}
\label{fig:kane_viz}
\end{figure}

\begin{figure}[bp!]
\centering
\includegraphics[width=\textwidth]{mean-and-variance.pdf}
\caption{
The effect of varying $p$ between zero and one when the trial stops after
seven patients respond or 11 do not.
}
\label{fig:kane_viz}
\end{figure}

%\begin{figure}[bp!]
%\centering
%\subfloat[enrollment distribution]{\includegraphics[width=0.45\textwidth]{snb-first-plot.pdf}}
%\hfill
%\subfloat[mean and variance varying $p$]{\includegraphics[width=0.45\textwidth]{mean-and-variance.pdf}}
%\caption{
%The distribution of the sample size in a trial that stops after seven 
%patients respond to treatment
%or 11 patients do not. Panel (a) shows the distribution when
%$p = 0.2$. The mean and variance of the distribution varying $p$ 
%between zero and one are shown in panel (b).
%}
%\label{fig:exp-and-var}
%\end{figure}

The distribution of the number of trial enrollments is shown in 
Fig.~\ref{fig:kane_viz}. There is relatively little probability mass
for values of sample size equal to seven through 10 since $p$ is small and it 
is unlikely the treatment will succeed quickly.
%The probability mass is concentrated at the 
%$11^{\text{th}}$ and $12^{\text{th}}$ step since $p$ 
%is small and for most realizations reach 11 
%non-responders before reaching 2 responders.
Fig.~\ref{fig:kane_viz} shows the expected value and variance for the
number of trial enrollments varying $p$ between zero and one. When $p$ is
small then the treatment
is more likely to fail shortly after the $11^{\text{th}}$ enrollment.
When $p$ is large then the treatment is more likely to succeed and the 
number of enrollees approaches seven from above.

@@

Let $\,b_1, b_2, \ldots \,$ denote a sequence of independent, identically
distributed, Bernoulli random variables with $\mathbb{P}[b_i=1]=p$ and
$\mathbb{P}[b_i = 0] = 1-p$, for
probability parameter $0\leq p \leq 1$. In the clinical trial setting
$\,b_i = 1$ corresponds to a patient responding to treatment.

Let $s$ and $t$ be positive integers.  Define the SNB random
variable $Y$ as the smallest
integer value such that $\,\{b_1, \ldots , b_Y\}\,$ contains {\em either}
$\,s\,$ responders {\em or} $\,t\,$ non-responders. That is, the SNB 
distribution of $Y$ is the smallest integer such that either
$\sum_i^Y b_i = s$ or $\sum_i^Y 1-b_i = t$.

The distribution of $\,Y\,$ has support on integer values in the range
\begin{equation*}               
     \min(s,t) \leq \; Y \;\leq s+t-1  \label{range.y.eq}.
\end{equation*}
The probability mass function is
\begin{equation} \label{eqn:pmf}
\mathbb{P} [Y=k] = S(k, p, s) \ I_{\{s \leq k \leq s+t-1\}} + 
  S(k, 1-p, t) \ I_{\{ t \leq k \leq s+t-1 \}}
\end{equation}
where $I_{\{f\}}$ is the {\em indicator function}, taking the value 
of one if $f$ is true and zero otherwise, and
\begin{equation} \label{eqn:N}
S(k, p, s) = {k-1 \choose s-1} p^s (1-p)^{k-s} 
\end{equation}
is the negative binomial probability mass. 
The CDF of the negative binomial is commonly expressed in terms of the
{\em regularized incomplete beta function} \citep{Olver2010}. Using standard
notation, (\ref{eqn:N}) is written as $\mathcal{I}_{1-p}(t, s)$.

@@

The probability a given realization of $\mathbf{X}$ reaches $s$ at
the $k$th outcome is the probability that, at time $k-1$, there are $s-1$
successful outcomes and $k-s$ unsuccessful outcomes multiplied by
the probability of a final success at time $k$. This expression is given
in (\ref{eqn:N}). 
Similarly, the probability a given realization reaches $k-t$
is the probability that, at outcome $k-1$, there are $k-t$ successful outcomes
and $t-1$ unsuccessful outcomes multiplied by the probability of a final
unsuccessful outcome at time $k$. That is,
\begin{align*}
p {k-1 \choose s=1} p^{s-1} & (1-p)^{k-s} + (1-p) {k-1 \choose k-t} p^{k-t} (1-p)^{t-1} \\ 
& = S(k, p, s) + S(k, 1-p, t)
\end{align*}
as claimed in (\ref{eqn:pmf}).

@@

To show (\ref{eqn:pmf}) sums to one, define
%\begin{align} \label{eqn:sum_proof}
%R &= \sum_{k=s}^{s+t-1} S(k, p, s) + \sum_{k=t}^{s+t-1} S(k, 1-p, t) \\
%  &= \sum_{k=s}^{s+t-1} {k-1 \choose s-1} p^s (1-p)^{k-s} + \sum_{k=t}^{s+t-1} {k-1 \choose k-t} p^{k-t} (1-p)^t
%\end{align}
\begin{equation*} 
R = \sum_{k=s}^{s+t-1} S(k, p, s) + \sum_{k=t}^{s+t-1} S(k, 1-p, t).
\end{equation*}
If we substitute $i=k-s$ in the first summation and $j=k-t$ in the second then
$R$ can be written as the cumulative distribution function (CDF) of two
negative binomial distributions:
\begin{equation} \label{eqn:transformed_sum}
R = \sum_{i=0}^{t-1} {i+s-1 \choose i} p^s (1-p)^i \; + \;
\sum_{j=0}^{s-1} {j+t-1 \choose j} p^j (1-p)^t.
\end{equation}

The regularized incomplete beta function satisfies
$\mathcal{I}_p(s, t) = 1-\mathcal{I}_{1-p}(t, s)$ \citep{Uppuluri1967}. Then
\begin{align*}
R = \sum_{i=0}^{t-1} &{i+s-1 \choose i} p^s (1-p)^i +
\sum_{j=0}^{s-1}  {j+t-1 \choose j} p^j  (1-p)^t \\
   &= 1-\mathcal{I}_p(s, t) + 1 - \mathcal{I}_{1-p}(t, s) \\
   &= 1. 
\end{align*}
This completes the proof (\ref{eqn:pmf}) is the distribution of the
stopping time and is a valid probability mass function.
\end{proof}

@@

\begin{figure}[p!]
\begin{center}
\includegraphics[width=0.8\textwidth]{shapes.pdf}
\end{center}
\caption{Different shapes of the SNB distribution with parameters 
($p$, $s$, $t$), as given. Black areas indicate the mass contributed by reaching
$s$ responders before $t$ non-responders. Grey indicates
mass contributed by reaching $t$ non-responders first. \label{shapes.fig}}
\end{figure}

@@

The SNB generalizes both the minimum (riff-shuffle) and maximum negative
binomial distributions up to a translation of the support.
For the special case of $\,s=t,$ the distribution of $\,Y\,$ is the
riff-shuffle, or minimum negative binomial 
distribution~\citep{Uppuluri1967,Johnson2005}.
The maximum negative binomial \cite{Johnson2005,Zhang2000,Zelterman2005} is 
the smallest number of outcomes necessary to 
observe at least $s$ responders {\em and} $s$ non-responders. This is equivalent
to a translated version of the riff-shuffle.

@@

%There is a close connection between the tail probabilities of the SNB and the 
%binomial distributions.
%The probability of reaching the success endpoint in an 
%SNB($p$, $s$, $t$) random variable is 
%equal to the probability of at least $s$ successes in a binomial distributed 
%random variable with size $s+t-1$ and success probability $p$. 
%Likewise, the probability of reaching the failure endpoint is equal
%to the probability of at most $s-1$ successes in a binomial distribution. 
There is also an equivalence between the probability of reaching an endpoint 
in the SNB model and the tail probability of the binomial distribution.
Specifically, the probability that the number of responders is at least 
$s$ in the binomial model is the same as the probability the treatment succeeds 
(reaches $s$) in the SNB model.
\begin{prop} \label{binomial_tail}
Let $Y$ be distributed as SNB($p$, $s$, $t$) and let $X_Y$ correspond
to the number of responders at the end of the trial. Let 
$B$ be distributed binomial with index parameter $n=s+t-1$ and response 
probability $p$. Then
\begin{equation}
\mathbb{P}[B \geq s] = \mathbb{P} [X_Y = s].
\end{equation}
\end{prop}
\begin{proof}
The binomial tail probability is
\begin{equation*}
\mathbb{P}[B \geq s] = 1 - \mathcal{I}_{1-p}(s, t)
\end{equation*}
The corresponding success probability is
\begin{equation} \label{eqn:success}
\mathbb{P} [X_Y = s] 
  = \sum_{k=s}^{s+t-1} {k-1 \choose s-1} p^s (1-p)^{k-s}.
\end{equation}
Let $i=k-s$. Since
\begin{equation*}
{i+s-1 \choose s-1} = {i+s-1 \choose i},
\end{equation*}
the summation in (\ref{eqn:success}) can be rewritten as
\begin{align*}
\mathbb{P} [X_Y = s] &= \sum_{i=0}^{t-1} {i+s-1 \choose i} p^s (1-p)^i\\
  &= 1 - \mathcal{I}_{1-p}(t, s)
\end{align*}
completing the proof.
\end{proof}

@@

The moment generating function for the SNB is calculated in a manner similar to 
that of two negative binomial distributions. 
\begin{prop} Let $Y$ be distributed SNB with parameters $p$, $s$, and $t$.
Then the moment generating function (MGF) of $Y$ is
\begin{equation} \label{eqn:mgf}
\mathbb{E}~e^{xY} = \left(\frac{p e^x}{1 - qe^x}\right)^s 
  \mathcal{I}_{1-qe^x} (s, t) + \left(\frac{qe^x}{1-pe^x}\right)^t 
  \mathcal{I}_{1-pe^x}(t, s)
\end{equation}
for $q = 1-p$ and is defined for 
$x < \min \left\{\log(1/p), \log(1/q) \right\}$.
\end{prop}
\begin{proof}
The MGF of the SNB is:
\begin{equation*}
\mathbb{E}~e^{xY} = \sum_{k=s}^{s+t-1} {k-1 \choose s-1} p^s q^{k-s} e^{kx} 
  + \sum_{k=t}^{s+t-1} {k-1 \choose t-1} p^{k-t} q^t e^{kx}
\end{equation*}
and can be rewritten as:
\begin{equation} \label{eqn:first_sum}
\mathbb{E}~e^{xY} = \sum_{k=s}^{s+t-1}{k-1 \choose s-1} (pe^x)^{s} (qe^x)^{k-s} 
  + \sum_{k=t}^{s+t-1}{k-1 \choose t-1} (qe^x)^t (pe^x)^{k-t}.
\end{equation}
The first summation in (\ref{eqn:first_sum}) satisfies
\begin{align*}
\sum_{k=s}^{s+t-1}{k-1 \choose s-1} (pe)^{sx} (qe^x)^{k-s} &= 
  \left(\frac{pe^x}{1 - qe^x}\right)^s \ \ \sum_{k=s}^{s+t-1} {k-1 \choose s-1} 
    (qe^x)^{k-s} (1-qe^x)^s \\
  &= \left(\frac{pe^x}{1 - qe^x}\right)^s \mathcal{I}_{1-qe^x}(s, t).
\end{align*}
Since the $p$ parameter in $\mathcal{I}_p$ has support on zero 
to one, we have $0 \leq pe^x < 1$. This implies $x < -\log(p)$.
A similar expression can be derived for the second summation in 
(\ref{eqn:first_sum}) and results in
the constraint $x < -\log(1-p)$.
\end{proof}

Depending on the parameter values the SNB can be approximated by 
the geometric, normal, gamma, and Poisson
distributions.

\begin{prop}
The MGF of the SNB converges to that of the negative binomial when either
$s$ or $t$ gets large. That is
\begin{equation*}
lim_{t\to\infty} \ \mathbb{E} e^{xY} = \left( \frac{pe^x}{1-qe^x} \right)^s
\end{equation*}
as. The analogous result holds when $s \rightarrow \infty$.
\end{prop}
\begin{proof}
The second incomplete beta function in (\ref{eqn:mgf}) can be written
in terms of a cumulative binomial distribution
\begin{equation*}
\mathcal{I}_{1-pe^x}(t, s) = \mathbb{P}\left[ B \leq s-1 \right]
\end{equation*}
where $B$ is distributed as
Binomial($t-k$, $pe^x$). From Chebychev's inequality %\cite{Hoeffding1963}
it follows that
\begin{equation} \label{eqn:hoeffding}
\mathbb{P}\left[ B \leq s-1 \right] \leq 
  \frac{ (t-k) pe^x (1-pe^x) }{ \left(s - (t-k)pe^x\right)^2 }
\end{equation}
As $t$ gets large $\mathcal{I}_{1-pe^x}(t, s)$ tends to zero
and $\mathcal{I}_{1-qe^x}(s, t)$ approaches 
one. The proof follows by realizing 
\begin{equation*}
0 < \frac{qe^x}{1-pe^x} < 1
\end{equation*}
over the support of $x$.
\end{proof}

When $s=1$ and $t$ is large the SNB's MGF is approximately 
the same as that of the
geometric distribution. The negative
binomial can therefore be seen as a sum of i.i.d. geometric distributions.
For an appropriately large number of distributions, the central limit theorem
yields a normal approximation.

Drawing connections to the gamma and Poisson distributions can also be shown.
A connection to the gamma distribution well-studied problem in the 
literature (see \cite{Best1974,Ord1968,Guenther1972} for examples). 
A connection to the Poisson appears in \cite{Anscombe1950} where
it is shown that when the mean of a Poisson is proportional
to a chi-square distribution with $2k$ degrees of freedom
then the negative binomial is obtained. 
Both of these approximations work by equating cumulants and then showing 
that differences between the cumulant generating
functions converge to zero.

The lower-half normal distribution can be approximated by setting $s=t$
for appropriately large $s$ and $t$ and $p = 0.5$. In this
case, the SNB can be viewed as identical, negative binomial distributions
approximating a normal and truncated at the median.

@@

Let us consider the Bayesian setting where the rate parameter is distributed 
as Beta($\alpha$, $\beta$) and denoted $P$.
The posterior distribution of $P$ is proportional to the 
likelihood, given by the function
%f_P(p, \alpha, \beta ) f_{Y|P}(p, k, s, t) &= 
\begin{align} \label{eqn:ptl}
f_{P|Y}(p, k, s, t, & \alpha, \beta) \propto
  \frac{ {k-1 \choose s-1} }{B(\alpha, \beta)} p^{\alpha +s -1} 
    (1-p)^{k+\beta-s-1} \\
  & + \frac{ {k-1 \choose t-1} }{B(\alpha, \beta)} p^{k+\alpha -t -1} 
    (1-p)^{\beta+t-1} \nonumber
\end{align}
where $0 \leq p \leq 1$, $s \leq k \leq s+k-1$, and $t \leq k \leq s+k-1$.
